\chapter{List of Publications}
\label{AppendixPubs}

\begin{enumerate}

  \item
  {\bf Michael H. Goerz}, Tommaso Calarco, and Christiane P. Koch. \\
  {\it The quantum speed limit of optimal controlled phasegates for trapped neutral atoms} \\
  J. Phys. B: At. Mol. Opt. Phys. {\bf 44}, 154011 (2011). \\
  Special issue on quantum control theory for coherence and information dynamics.
  \label{GoerzJPB2011}

  \item
  Michał Tomza, {\bf Michael H. Goerz}, Monica Musiał, Robert Moszyński, and Christiane P. Koch. \\
  {\it Optimal production of ultracold ground-state molecules: Stabilization employing potentials with ion-pair character and strong spin-orbit coupling.}  \\
  Phys. Rev. A {\bf 86}, 043424 (2012).
  \label{TomzaPRA2012}

  \item
  {\bf Michael H. Goerz}, Daniel M. Reich, and Christiane P. Koch. \\
  {\it Optimal control theory for a unitary operation under dissipative evolution.} \\
  New J. Phys. {\bf 16}, 055012 (2014).  \\
  Special issue on coherent control of complex quantum systems.
  \label{GoerzNJP2014}

  \newpage
  \item
  Henning A. Fürst, {\bf Michael H. Goerz}, Ulrich G. Poschinger, Michael Murphy, Simone Montangero, Tommaso Calarco, Ferdinand Schmidt-Kaler, Kilian Singer, and Christiane P. Koch. \\
  {\it Controlling the transport of an ion: Classical and quantum mechanical solutions.} \\
  New J. Phys. {\bf 16}, 075007 (2014). \\
  Special issue on coherent control of complex quantum systems.
  \label{FuerstNJP2014}

  \item
  Georg Jäger, Daniel M. Reich, {\bf Michael H. Goerz}, Christiane P. Koch, and Ulrich Hohenester. \\
  {\it Optimal quantum control of Bose-Einstein condensates in magnetic microtraps: Comparison of GRAPE and Krotov optimization schemes.} \\
  Phys. Rev. A {\bf 90}, 033628 (2014).
  \label{JaegerPRA2014}

  \item
  {\bf Michael H. Goerz}, Eli J. Halperin, Jon M. Aytac, Christiane P. Koch, and K.  Birgitta Whaley. \\
  {\it Robustness of high-fidelity Rydberg gates with single-site addressability.} \\
  Phys. Rev. A {\bf 90}, 032329 (2014). \\
  Editor's suggestion.
  \label{GoerzPRA2014}

  \item
  Paul Watts, Ji\v{r}\'i Vala, Matthias M. M\"uller, Tommaso Calarco, K.\ Birgitta Whaley, Daniel M.\ Reich, {\bf Michael H. Goerz}, and Christiane P.\ Koch. \\
  {\it Optimizing for an arbitrary perfect entangler. I.~Functionals.} \\
  arXiv:1412.7347. Accepted in Phys. Rev. A.
  \label{PE1Paper}

  \item
  {\bf Michael H. Goerz}, Giulia Gualdi, Daniel M. Reich, Christiane P. Koch, Felix Motzoi, K. Brigitta Whaley, Ji\v{r}\'i Vala, Matthias M. M\"uller, Simone Montangero, and Tommaso Calarco. \\
  {\it Optimizing for an arbitrary perfect entangler. II.~Application.} \\
  arXiv:1412.7350. Accepted in Phys. Rev. A.
  \label{PE2Paper}

  \item
  {\bf Michael H. Goerz}, K. Birgitta Whaley, and Christiane P. Koch\\
  {\it Hybrid Optimization Schemes for Quantum Control.} \\
  arXiv:1505.05331. Submitted to EPJ Quantum Technol.
  \label{HybridPaper}

\end{enumerate}

\noindent
Chapters~\ref{chap:robust},~\ref{chap:pe},~\ref{chap:3states} of this thesis are
based on the publications (\ref{GoerzNJP2014}), (\ref{PE1Paper}) and
(\ref{PE2Paper}), and(\ref{GoerzPRA2014}), respectively.
In parts of the chapters, text has been been taken verbatim from the
corresponding publications, but only if it was the author's
original work.
